Reproducible development environments with rix

### 2023-12-14 paper.qmd
started nix

As we go down this list, we get closer to the gold standard of a perfectly
reproducible study.

However, reaching this gold standard is quite costly: one needs to learn a tool
to deal with package versions, then a build automation tool such as
[make]{.proglang} and ideally [Docker]{.proglang} should be added to the list.
[Docker]{.proglang} is a tool that makes it possible to run arbitrary code in a
completely controlled and isolated environment as a way to capture the complete
underlying system libraries. Simply put, a `Dockerfile` provides a complete
description of the computational environment as well as the required instructions
to build the complete project. One can then use [Docker]{.proglang} to build
a so-called *image* which contains all the required software and instructions
to build the project. From this image, it is then possible to run a *container*
which actually runs the analysis.

### Prior art {#subsec-prior}

>>>>

However, reaching this gold standard is quite costly: one needs to learn a tool
to deal with package versions, then [Docker]{.proglang} for all the other
software, including the programming language that was used. Ideally, a build
automation tool such as [make]{.proglang} should be added to the list. It should
also be noted that these requirements are not limited to the [R]{.proglang}
programming language. A very similar approach should be taken if one uses 
[Python]{.proglang} for statistical analysis instead.

As an alternative to the above approach, I will present the [Nix]{.proglang}
package manager, which available for all major operating system and focuses on
installing and building software in a reproducible manner. To make
[Nix]{.proglang} more accessible to [R]{.proglang} programmers, I developed the
[rix]{.pkg} which I will also present in this article.

## The [Nix]{.proglang} package manager

[Nix]{.proglang} is a package manager that can be used to install and build
software in a completely reproducible manner. With [Nix]{.proglang}
it is possible to install [R]{.proglang} and [R]{.proglang} packages as well
(whether these come from CRAN or Bioconductor)

is that its repository of software is huge. As of writing, it contains more than
80.000 packages, and the entirety of CRAN and Bioconductor is available through
Nix’s repositories. This means that using Nix, it is possible to install not
only R, but also all the packages required for your project. The obvious
question is why use Nix instead of simply installing R and R packages as usual.
The answer is that Nix makes sure to install every dependency of any package, up
to required system libraries. For example, the `{xlsx}` package requires the
Java programming language to be installed on your computer to successfully
install. This can be difficult to achieve, and `{xlsx}` bullied many R
developers throughout the years (especially those using a Linux distribution,
`sudo R CMD javareconf` still plagues my nightmares). But with Nix, it suffices
to declare that we want the `{xlsx}` package for our project, and Nix figures
out automatically that Java is required and installs and configures it. It all
just happens without any required intervention from the user. The second
advantage of Nix is that it is possible to pin a certain revision of the Nix
packages’ repository (called nixpkgs) for our project. Pinning a revision
ensures that every package that Nix installs will always be at exactly the same
versions, regardless of when in the future the packages get installed.

With Nix, it is essentially possible to replace {renv} and Docker combined, or
if you’re using mainly Python, you can replace `conda` or `requirements.txt`
files. If you need other tools or languages like Python or Julia, this can also
be done easily. Nix is available for Linux, macOS and Windows (via WSL2).
Important remark: since using Nix on Windows must go through WSL, when we refer
to "Linux" in the context of Nix, this includes Windows by default as well. It
is also possible to build multi-language environments, containing R and Python,
a LaTeX distribution and packages and so on.

## The Nix programming language

Nix is not just useful because it is possible to install many packages and even
install older packages, but also because it comes with a complete functional
programming language. This programming language is used to write *expressions*,
and these expressions in turn are used to build software. Essentially, when you
install a package using Nix, an expression gets downloaded from the Nix package
repository (more on that in the next section), and it gets evaluated by the Nix
package manager. This expression contains a so-called *derivation*. A derivation
defines a build: some inputs, some commands, and then an output. Most of the
time, a derivation downloads source code, builds the software from the source
and then outputs a compiled binary. Derivations are extremely flexible, and you
could write a derivation to build a complete environment and then build a
complete reproducible pipeline. The output could be any of the discussed data
products.

Learning the Nix programming language is a good idea if you want to contribute
to the Nix package repository, but you might not have to learn it in-depth if
you simply wish to use it to build reproducible environments, as we will learn
now. If you wish to learn about the programming language, I highly recommend [a
tour of
Nix](https://nixcloud.io/tour/?id=introduction/nix)^[https://nixcloud.io/tour/?id=introduction/nix].

## The Nix package repository

So, there’s the Nix package manager, the Nix programming language and the Nix
package repository (henceforth nixpkgs). To look for packages click
[here](https://search.nixos.org/packages)^[https://search.nixos.org/packages].
The source code of all the packages (so the whole set of Nix expressions) can be
found on [this Github
repository](https://github.com/NixOS/nixpkgs)^[https://github.com/NixOS/nixpkgs].
For example,
[here](https://github.com/NixOS/nixpkgs/blob/master/pkgs/development/libraries/quarto/default.nix)^[https://github.com/NixOS/nixpkgs/blob/master/pkgs/development/libraries/quarto/default.nix]
is the Nix expression that contains the derivation to build `quarto`. As you can
see, the derivation uses the the pre-built Quarto binaries instead of building
it from source. Adding packages to nixpkgs (or updating them) can be done by
opening pull requests. For example,
[here](https://github.com/NixOS/nixpkgs/pull/259443)^[https://github.com/NixOS/nixpkgs/pull/259443]
is a pull request to make Quarto available to all platforms (before this PR
Quarto was only available for Linux). PRs get reviewed and approuved by
maintainers that also have the right to merge the PR into master. Once merged,
the new or updated package is available for download. Because nixpkgs is a
“just” Github repository, it is possible to use a specific commit hash to
install the packages as they were at a specific point in time. For example, if
you use this commit, 7c9cc5a6e, you’ll get the packages as of the 19th of
October 2023, but if you used this one instead: 976fa3369, you’ll get packages
from the 19th of August 2023. Using specific hashes is called "pinning" and you
can read more about it [here](https://nixos.wiki/wiki/FAQ/Pinning_Nixpkgs). We
will make extensive use of pinning.
